Parallel Acceleration of Fixed Complexity Sphere Decoder for Large MIMO Uplink System based on GPU

### 2014-10-30 parallel_acceleration_of_FCSD_based_on_cuda/draft.tex
revised up to page 13

In this section we present the parallel implementation of fixed complexity sphere decoder by GPU, first from (\ref{formula 1}), the maximum likelihood (ML) solution is given by:

@@

s_{ML}=arg\min_{s\in \mathbb{O}^{N_{t}}}||y-\mathbf{H}s||^{2} \label{formula 2}

@@

we consider the array $N_{t}=N_{r}$, performing QR decomposition to channel propagation matrix

@@

where $\mathbf{Q}\in \mathbf{C}^{N_{t}\times N_{t}}$ is the unitary matrix, $\mathbf{R}\in \mathbf{C}^{N_{t}\times N_{t}}$, is the upper triangular matrix, the basing on \cite{matrix theory}, the (\ref{formula 2}) can be changed to

@@

s_{ML}&=&arg\min_{s\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}y-\mathbf{Q}^{H}\mathbf{H}s||^{2}\\
&=& arg\min_{s\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}y-\mathbf{R}s||^{2} \label{formula 3}

@@

Consider the unconstrained estimation of $s$, $\hat{s}=\mathbf{G}y$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$ denotes the zero forcing equalizer, basing on \ref{QR}, we have  

@@

\hat{s}=\mathbf{G}y=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}y=\mathbf{R}^{-1}\mathbf{Q}^{H}y 

@@

Therefore $\mathbf{Q}^{H}y=\mathbf{R}\hat{s}$, the \ref{formula 3} can be changed to

@@

s_{ML}=arg\min_{s\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\hat{s}-s)||^{2} \label{formula 4}

@@

the Fixed Complexity Sphere Decoder works as a constant number of multiple path tree searching.

@@

\caption{Tree searching of QPSK FCSD for 8x8 MIMO system}

@@

As shown in Figure \ref{figure4}, just like SD the FCSD also performs depth first searching, at the first $\rho$ level the FCSD search all the possible signal symbols $\widetilde{s}_{N_{t}} \widetilde{s}_{N_{t}-1}\dots \widetilde{s}_{N_{t}-\rho+1}$ in constellation $\mathbb{O}^{N_{t}}$, which is called Full Expansion(FE) level, the remain $N_{t}-\rho$ levels of symbols, the FCSD performs decision feed back equalization (DFE), from (\ref{formula 4}) the solution of FCSD can be given:

@@

\hat{s}_{FCSD}=arg\min_{s\in \mathbb{O}^{N_{t}}}\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s_{j})|^{2}  \label{formula 5}

@@

$i\in [1,2\dots N_{t}]$, $r_{ij}$ denotes the element at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$, this process is called Single Expansion (FE).
thus the FCSD solution $s_{FCSD}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:

@@

\in \mathbb{O}^{\rho}      \quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\

@@

$\mathbb{Q}[]$ denotes the signal constellation based quantization. At the post processing stage, FCSD takes the symbol vector candidate with the minimum Euclidean metric $E_{metric}$:

@@

E_{metric}=||\mathbf{R}(\hat{s}-\widetilde{s})||^{2}

@@

The $\widetilde{s}$ denotes the symbol vector candidate, the total number of candidates is: 

@@

\subsection{CUDA-FCSD Pre-Prossing}
As shown in figure.4, each path of the FCSD searching has a serial execution nature, in order to avoid error propagation of decoder, FCSD ordering was applied to the propagation channel based on post processing signal to noise ratio\cite{Vblas}, which can be expressed as:

@@

$\varphi_{m}$ denotes the post processing signal to noise ratio of the $m$th data stream, $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$ denotes the diagonal elements of the inversion of $HHH$ matrix $\mathbf{H}^{H}\mathbf{H}$, thus when the SNR is determined, the post processing SNR is only determined by $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$, for the FE stage the robustness of detector performance is not influenced by the post processing SNR thus at FE stage the "weakest" data stream is detected first, that is, the data stream with the smaller $\varphi$ is detected earlier. At SE stage, because of the single searching path of the symbol, the performance is tightly related to the post processing SNR, therefore at SE stage the data streams with the larger post processing SNR is detected earlier in order to avoiding the error propagation, in conclusion the FCSD ordering works iteratively  and obeys the following rule:
At the $j$th step ($j=1,2\dots N_{t}-1$), $s_{p}$ denotes the symbol to be detected, where 

@@

arg \max_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad FE stage\\
arg \min_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad SE stage \\  \label{the ordering}

@@

$\mathbf{H}_{j}\in \mathbb{C}^{N_{r}\times (N_{t}-j+1)}$ denotes the renewed channel matrix, at the $j$th iteration with the column corresponding to previous detected data streams removed from $\mathbf{H}_{j-1}$.
the unconstrained estimation $\hat{s}$ in \ref{unconstrained estimation}  in is given by:
 \hat{s}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}y \label{shat}

Although the FCSD pre-processing has a serial nature and can not be vectorized, the computational consumptive operations \ref{the ordering} \ref{shat} can be implement by the CUDA Basic Linear Algebra Subprograms(cuBLAS), the matrix inverse(Gaussian Jordan Elimination) and cholesky factorization(Row Major Factorization) are also implemented by GPU, as shown in Table \ref{table 1}.
\begin{table}[htb]
\caption{Parallel Computation Operation of FCSD Pre-processing}
\centering
\begin{tabular}{|c|p{3cm}|p{3cm}|p{3cm}|p{3cm}|}
\hline
Operations & complex matrix-matrix multiplication & complex matrix-vector multiplication & matrix inverse & cholesky factorization \\
\hline
GPU subroutine & $\mathit{cublasCgemm}$  & $\mathit{cublasCgemv}$ & $\mathit{GJE}$ & $\mathit{chol}$\\
\hline
Formula   &  $\mathbf{H}^{H}\mathbf{H}$ & $(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}y$ & $\mathbf{H}_{j}^{H}\mathbf{H}_{j})^{-1} $ & $\mathbf{H}^{H}\mathbf{H}=\mathbf{R}^{H}\mathbf{R}$ 
\\
\hline
  \label{table 1}
\end{tabular}
\end{table}

>>>>

\paragraph{}In this section we present the parallel implementation of fixed complexity sphere decoder using GPU, From (\ref{formula 1}), the maximum likelihood (ML) detector for a MIMO system is specified by:

@@

s_{ML}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{y}-\mathbf{H}\mathbf{s}||^{2} \label{formula 2}

@@

We consider a MIMO system with  $N_{t}=N_{r}$. Performing the QR decomposition to the channel propagation matrix

@@

where $\mathbf{Q}\in \mathbf{C}^{N_{t}\times N_{t}}$ is an unitary matrix and $\mathbf{R}\in \mathbf{C}^{N_{t}\times N_{t}}$, is an upper triangular matrix, the basing on \cite{matrix theory}, we can write (\ref{formula 2}) as:

@@

\mathbf{s}_{ML}&=&arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{Q}^{H}\mathbf{H}\mathbf{s}||^{2}\\
&=& arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{Q}^{H}\mathbf{y}-\mathbf{R}\mathbf{s}||^{2} \label{formula 3}

@@

Consider the unconstrained estimation of $\mathbf{s}$, $\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}$, $\mathbf{G}=(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}$ based on \ref{QR}, we have  

@@

\mathbf{\hat{s}}=\mathbf{G}\mathbf{y}=(\mathbf{R}^{H}\mathbf{R})^{-1}\mathbf{R}^{H}\mathbf{Q}^{H}\mathbf{y}
=\mathbf{R}^{-1}\mathbf{Q}^{H}\mathbf{y}

@@

Therefore $\mathbf{Q}^{H}\mathbf{y}=\mathbf{R}\mathbf{\hat{s}}$, and \ref{formula 3} can be written as

@@

\mathbf{s_{ML}}=arg\min_{\mathbf{s}\in \mathbb{O}^{N_{t}}}||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{s})||^{2} \label{formula 4}

@@

@@

\caption{Tree searching of 4-QAM FCSD for 8x8 MIMO system}

@@

\paragraph{}Just like SD the FCSD also performs depth first searching, at the first $\rho$ node levels the FCSD search all the possible signal symbols in constellation $\mathbb{O}^{N_{t}}$, which is called Full Expansion(FE) stage, therefore for the symbol nodes $\widetilde{s}_{N_{t}} \widetilde{s}_{N_{t}-1}\dots \widetilde{s}_{N_{t}-\rho+1}$, the FCSD has $M$ branches expansion at each symbol node,  as shown in Figure \ref{figure4}, there are $\rho=2$ FE nodes and each level has $M=4$ branches. (\ref{formula 4}) can be written in the form: 

@@

\hat{s}_{FCSD}=arg\min_{s\in \mathbb{O}^{N_{t}}}\{\sum_{i=1}^{N_{t}}|r_{ii}(\hat{s}_{i}-s_{i})+\sum_{j=i+1}^{N_{t}}r_{ij}(\hat{s}_{j}-s_{j})|^{2}\}  \label{formula 5}

@@

$i\in [1,2\dots N_{t}]$, $r_{ij}$ denotes the component at the $i$th row and $j$th column of the upper triangular matrix $\mathbf{R}$,  based on \ref{formula 5} at the remaining $N_{t}-\rho$ levels of symbols, the FCSD performs decision feed back, therefore there are only $1$ branch expansion for each symbol node at this stage, this stage is called Single Expansion (SE). 
thus the total number of the branches is fixed (For example in Figure (\ref{figure4}) the total branch number is $4^{2}$), the FCSD works as a constant number of multiple path tree searching algorithm.
In conclusion the FCSD solution $s_{FCSD}=[s_{1},s_{2}\dots s_{N_{t}}]$ can be expressed as:

@@

\Omega\in  \mathbb{O}^{\rho}      \quad i=N_{t},N_{t}-1,\dots N_{t}-\rho+1\\

@@

$\mathbb{Q}[\cdot]$ denotes the signal constellation quantization. At the post processing stage, FCSD compares the Euclidean metric of all the symbol vector candidates $E_{metric}$, which, according to \ref{formula 4}, is defined as:

@@

E_{metric}=||\mathbf{R}(\mathbf{\hat{s}}-\mathbf{\widetilde{s}})||^{2}

@@

Where the $\mathbf{\widetilde{s}}$ denotes the symbol vector candidate. The total number of candidates is: 

@@

The symbol vector candidate with the minimum $E_{metric}$ is chosen as the solution.
\subsection{CUDA-FCSD Preprossing}
\paragraph{}As shown in Figure \ref{figure4}, each path of the FCSD searching has a serial execution nature. In order to avoid error propagation, FCSD ordering was applied to the propagation channel based on post processing signal to noise ratio\cite{wolniansky1998v}:

@@

where $\varphi_{m}$ denotes the post processing signal to noise ratio of the $m$th data stream, and $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$ denotes the diagonal elements of the inversion of $\mathbf{H}^{H}\mathbf{H}$. When the SNR is determined, the post processing SNR is only determined by $(\mathbf{H}^{H}\mathbf{H})_{mm}^{-1}$, From the heuristic in \cite{barbero2008fixing} for the FE stage the robustness of detector performance is not influenced by the post processing SNR. Thus at FE stage the "weakest" data stream, that is, the data stream with the smallest $\varphi$, is detected first. At SE stage, because of the single searching path of the symbol, the performance is tightly related to the post processing SNR. Therefore at SE stage the data streams with the larger post processing SNR is detected earlier in order to avoiding error propagation, In conclusion the FCSD ordering works iteratively  and obeys the following rule:
at the $j$th step ($j=1,2\dots N_{t}-1$), $s_{p}$ denotes the symbol to be detected, where 

@@

arg \max_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad FE\quad stage\\
arg \min_{k}(\mathbf{H}_{j}^{H}\mathbf{H}_{j})_{kk}^{-1}\quad SE\quad stage   \label{the ordering}

@@

and $\mathbf{H}_{j}\in \mathbb{C}^{N_{r}\times (N_{t}-j+1)}$ denotes the renewed channel matrix. At the $j$th iteration with the column corresponding to previous detected data streams removed from $\mathbf{H}_{j-1}$.
the unconstrained estimation $\mathbf{\hat{s}}$  is given in (\ref{unconstrained estimation}):

\paragraph{}Although the FCSD pre-processing has a serial nature and can not be vectorized, the computational consumptive operations \ref{the ordering} \ref{shat} can be implement by the CUDA Basic Linear Algebra Subprograms(cuBLAS), the matrix inverse(Gaussian Jordan Elimination) and cholesky factorization(Row Major Factorization) are also implemented by GPU, as shown in Table \ref{table 1}.
\begin{table}[htb]
\caption{Parallel Computation Operation of FCSD Pre-processing}
\centering
\begin{tabular}{|c|p{3cm}|p{3cm}|p{3cm}|p{3cm}|}
\hline
Operations & complex matrix-matrix multiplication & complex matrix-vector multiplication & matrix inverse & cholesky factorization \\
\hline
GPU subroutine & $\mathit{cublasCgemm}$  & $\mathit{cublasCgemv}$ & $\mathit{GJE}$ & $\mathit{chol}$\\
\hline
Formula   &  $\mathbf{H}^{H}\mathbf{H}$ & $(\mathbf{H}^{H}\mathbf{H})^{-1}\mathbf{H}^{H}y$ & $\mathbf{H}_{j}^{H}\mathbf{H}_{j})^{-1} $ & $\mathbf{H}^{H}\mathbf{H}=\mathbf{R}^{H}\mathbf{R}$ 
\\
\hline
  \label{table 1}
\end{tabular}
\end{table}
